Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2016-07-18 ms/ms.Rnw
Draft dicussion

\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and plot-level effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates. This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.

@@

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.

Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does mortality vary substantially between censuses?
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality strongly varies over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees dying between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{\beta X_{i}} + \gamma)\delta$) and used stem diameter growth rather than basal-area growth, significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e. $(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig. \ref{fig2}A). Surprisingly, we found that the growth-dependent only function did not improve predictive capacity over the baseline only function. Furthermore, we found that while the hazard function that incorporates both baseline and growth-dependent terms significantly improved model accuracy, this was only apparent if dbh growth was used.

@@

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). The most parsimonious combination that provided the largest gain in predictive accuracy (i.e. lowest logarithmic loss) occurred when only the baseline hazard term ($\gamma$) was varied as a function of species wood density ($\rho$; Fig. \ref{fig2} B). We found that wood density was negatively correlated with species baseline mortality rate (Fig. \ref{fig3}). This meant that species with low wood density had on average higher growth-independent mortality rates ((Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to high wood density species. While, wood density did improve model accuracy it only accounted for XX\% of the total gain in predictive capacity that could be obtained by accounting for species differences.

\subsection*{Proportion of variation explained}
Predicted mortality rates were most variable between species accounting for approximately XX\% followed by between census periods (accounting for XX\%). Wood density, a species level trait, accounted for approximately XX\% of the total species variability, suggesting that other species differences contribute to the ramaining XX\% unexplained species variation.

As expected, the growth-dependent hazard terms (i.e. the part of the model allowed to vary by individual growth) accounted for XX\% of the overall variation in mortality rates while the baseline (growth-independent) hazard accounted for XX\%.

@@

\subsection*{Mortality rates, species light requirements and species maximum dbh}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species that predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth stochastic chance and required faster growth rates for a given mortality rate relative to a species that readily recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

\begin{figure}[h!]
\includegraphics{figures/fig5.pdf}
\caption{Proportion of predicted variation explained.
A) Variation explained by wood density (grey) as well as species and census random effects.
B) Variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
\label{fig5}
\end{figure}

\section*{Discussion}

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

\begin{figure}[h!]
\includegraphics{figures/fig6.pdf}
\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (A,C,E,G) and species maximum DBH (B,D,F,H). Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study. Blue line shows the average trendline as determined from standard linear regression}
\label{fig6}
\end{figure}

Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.

- General overview of results:
As expected, the most biologically reasonable model; the model with both a baseline hazard and growth-dependent hazard was the best predictor of tropical tree mortality, indicating that both growth dependent and independent hazards are needed to effectively model tree mortality.

- Furthermore, we found that DBH was by far the better growth measure for predicting plant mortality.
	- Why DBH? Could it be that DBH implicitly punishes smaller individuals relative to larger individuals. That is, small plants can only gain a small amount of DBH growth relative to a larger plant. As such, DBH growth implicitly captures the process whereby small individual plants are more prone to competitive effects as well as random breakage events caused by trampling, herbivory etc.

- As expected, when allowing parameters to vary by species we found a significant improvement in model predictive capacity.
		- This supports previous research that shows that species vary substantially in mortality rates.

- When examining this interspecific variability we found that the effect of wood density improved model predictive capacity when it was included on the base-line hazard, with only marginal improvements when included on all three parameters.

- Furthermore, we found that wood density explained XX\% of the total species variation for the baseline hazard.

- Conclusion, species with higher wood density had lower baseline hazards supporting previous work that suggests that these species are less prone to stochastic events that involve breakage etc.

@@

The most complicated function is
\begin{equation}
\lambda_i = \big(\underbrace{\alpha \, \exp\left(-\beta \, X_i \right)}_{\text{Growth-dependent}} +  \underbrace{\gamma}_{\text{Baseline}}\big) \underbrace{\delta_t}_{\text{Census-effect}}.
 \label{add_haz}
\end{equation}

>>>>

% Enable cross referencing to Evidence document
\usepackage{xr}
\externaldocument{ms-suppinfo}
\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008). This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.

@@

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.

Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
  \item Does mortality vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality strongly varies over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e. $(\alpha e^{\beta X_{i}})\delta$) hazard functions, irrespective of growth rate (Fig. \ref{fig2}A). However, predictive accuracy was higher for stem diameter relative to basal-area growth. Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.

@@

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). The most parsimonious combination that provided the largest gain in predictive accuracy (i.e. lowest logarithmic loss) occurred when only the baseline hazard term ($\gamma$) was varied as a function of species wood density ($\rho$; Fig. \ref{fig2} B). We found that wood density was negatively correlated with species baseline mortality rate (Fig. \ref{fig3}). This meant that species with low wood density and fast individual growths had on average higher mortality rates ((Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to high wood density species. While, wood density did improve model accuracy it only accounted for \Sexpr{pred_gain_wd(logloss_summaries)}\% of the total gain in predictive capacity that could be obtained by accounting for species differences.

\subsection*{Proportion of variation explained}
Approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates captured by our model was due to species differences, \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density. Census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
As expected, the growth-dependent hazard terms (i.e. the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates.

@@

\subsection*{Mortality rates, maximum species size and species light requirements}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species which predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

\begin{figure}[h!]
\includegraphics{figures/fig5.pdf}
\caption{Proportion of predicted variation explained.
A) Predicted variation explained by wood density (grey) as well as species and census random effects.
B) Predicted variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
\label{fig5}
\end{figure}

\section*{Discussion}
Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal varability. Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, only explained a small proportion of the overall inter-specific variation.

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards (Moorcroft, Sortie,McDowell 2011). We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed. Like many other studies (Denslow JS (1987) King DA, Davies SJ, Supardi MNN, Tan S (2005), Ruger 2011), our analyses highlight the importance of light competition as a dominant factor influence tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig 5), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.

Incorporating the effect of wood density on mortality rates also improved predictive performance. Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality. Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts. This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}. More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.

While wood density effects are now being incorporated in mortality algorithms of many vegetation models (REFS), our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation. This means that such models are likely to severely underestimate the true varability in mortality rates. Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics \cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.

\begin{figure}[h!]
\includegraphics{figures/fig6.pdf}
\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (A,C,E,G) and species maximum DBH (B,D,F,H). Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study. Blue line shows the average trendline as determined from standard linear regression. Gap index is a binary species level index of the light required for recruitment. Here low numbers indicate the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
\label{fig6}
\end{figure}

The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates. Poorter et al. (2006) suggested that a species light demand and maximum size (a correlate to species longevity; Lieberman et al 1985) are two major axes of inter-specific variation in tropical rainforests. While we did not explicitly model mortality as a function of these traits, our posthoc correlations (FIG) revealed that some inter-specific variation was indeed correlated with light demand, but not maximum size. We found that light demanding species (i.e. those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions.

We recommend research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability. Researchers, however, should be mindful that the inclusion of additional traits comes at the cost of added complexity, increased uncertainty (i.e. more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems. Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality. For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture (REF).

Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally (McDowell 2011, Get a Zepple paper). Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, based on anedoctal evidence, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a 1998 El-Nino (REF). Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth (Chave's biomass paper) will compare.

Here we showcase a new framework for modelling tropical tree mortality that attempts to unifies empirical evidence from within and between species studies. This framework also provides an approach for partitioning mortality rates into growth-dependent and growth-independent hazards. Our findings reveal that while wood density is an important trait affecting mortality rates, we are still only capturing a fraction of the overall species variability in mortality rates.

@@
